Cell Decompositions for Rank Two Quiver Grassmannians

### 2017-11-20 Grassmannian_cells.tex
Consequences for F-polynomials and cell decompositions

\newcommand{\bfe}{\mathbf{e}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\bfg}{\mathbf{g}}
\newcommand{\bfi}{\mathbf{i}}
\newcommand{\bfh}{\mathbf{h}}
\newcommand{\tbfe}{{\tilde\bfe}}
\newcommand{\tbff}{{\tilde\bff}}
\newcommand{\tbfg}{{\tilde\bfg}}
\newcommand{\tbfh}{{\tilde\bfh}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\cG}{\mathcal{G}}
\newcommand{\cP}{\mathcal{P}}
\newcommand{\ui}{\underline i}
\newcommand{\uj}{\underline j}
\newcommand\udim{{\underline{\dim}\, }}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Rep}{\mathrm{Rep}}
\newcommand{\rep}{\operatorname{rep}}
\newcommand{\Gr}{\mathrm{Gr}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\kk}{\Bbbk}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\Ext}{\operatorname{Ext}}
\newcommand{\Hom}{\operatorname{Hom}}
\newcommand{\im}{\operatorname{im}}
\newcommand{\codim}{\operatorname{codim}}
\newcommand{\into}{\hookrightarrow}
\newcommand{\onto}{\to\!\!\!\!\!\to}
\newcommand{\ses}[3]{0\rightarrow #1\rightarrow #2\rightarrow#3\rightarrow 0}
\newcommand{\Sc}[2]{\langle #1,#2\rangle}
\newcommand{\spn}{\operatorname{span}}
\setcounter{MaxMatrixCols}{20}

@@

\begin{theorem}Let $X$ be a representation which can be lifted to $\hat Q$ and choose $d:\mathrm{supp}(X)\to \ZZ$ such that it satisfies the degree condition for $X$. Then the torus action of $T$ on $\bigoplus_{q\in Q_0}X_q$ induced by $t.x_{(q,\chi)}=t^{d(q,\chi)}x_{(q,\chi)}$ for $x_{(q,\chi)}\in X_{(q,\chi)}$ induces
 a torus action on $Gr_\bfe^Q(X)$ such that
  $$Gr^Q_\bfe(X)^T\cong\bigsqcup_{[\hat\bfe]} Gr^{\hat Q}_{\hat\bfe}(\hat X)$$ where $\hat\bfe$ runs through all equivalence classes of dimension vectors compatible with $\bfe$.
\end{theorem}
\begin{proof}
A representation $U\in\Gr_{\bfe}(X)$ is a torus fixed point if and only if $t.U=U $ for all $t\in\C^\ast$, i.e. $t.U_q=U_q$ for all $q\in Q_0$, $t\in\C^\ast$. Thus, apart from being a subrepresentation of $X$, it is a fixed point of the torus actions on the usual Grassmannians $\Gr_{\bfe_q}(X_q)$ induced by the decompositions $X_q=\bigoplus_{\chi\in A_Q} X_{q,\chi}$. A subrepresentation $U\subset X$ can be lifted to the universal abelian covering if and only if
\[X_q\cap U_q=\bigoplus_{\chi\in A_Q} X_{(q,\chi)}\cap U_{(q,\chi)}=\bigoplus_{\chi\in A_Q}U_{(q,\chi)}.\]
Thus we can apply Lemma \ref{usualGrass} to obtain the claim.
\end{proof}
The next step is to iterate the torus action keeping in mind the following idea: every representation, which lifts to the universal covering quiver, also lifts to the universal abelian covering quiver and also to the iterated universal abelian covering quivers, i.e. to $\hat Q^{(k)}:=\widehat{\hat Q^{(k-1)}}$ with $\hat Q=\hat Q^{(1)}$. Now it is straightforward to check that there exist natural surjective morphisms $f_k:\tilde Q\to \hat Q^{(k)}$ which become injective on finite subquivers if $k\gg 0$, see also \cite[Section 3.4]{wei}. Since the support of $X$ is finite as a representation of $\tilde Q$, we can find $k\geq 1$ such that the full subquiver with vertices $\mathrm{supp}(X)\subseteq \hat Q^{(k)}_0$ is a tree. Thus, if $\hat X^{(k)}$ is the lift of $X$ to $\hat Q^{(k)}$, there exists a torus action on the vector spaces $\hat X^{(k)}_q$ which induce a torus action on the quiver Grassmannian $Gr_{\bfe}^{\hat Q^{(k)}}(\hat X^{(k)})$. If we denote the iterated torus fixed points by $Gr^Q_\bfe(X)^{T^{(k)}}$. We obtain:
\begin{corollary}
Let $X$ be a representation which can be lifted to $\tilde Q$. Then there exists an iterated torus action such that
  $$Gr^Q_\bfe(X)^{T^{(k)}}\cong \bigsqcup_{[\hat\bfe^{(k-1)}]} Gr_{\hat\bfe^{(k-1)}}^{\hat Q^{(k-1)}}(\hat X^{(k-1)})^T\cong \bigsqcup_{[\hat\bfe^{(k)}]} Gr^{\hat Q}_{\hat\bfe^{(k)}}(\hat X^{(k)})\cong \bigsqcup_{[\tilde\bfe]} Gr^{\tilde Q}_{\tilde\bfe}(\tilde X)$$ where $\hat\bfe^{(k-1)},\,\hat\bfe^{(k)},\,\tilde\bfe$ run through all equivalence classes of dimension vectors compatible with $\bfe$.
\end{corollary}
%\begin{corollary}
%  affine bundles over $Gr^{\tilde Q}_{\tilde\bfe}(\tilde X)$ if $Gr^Q_\bfe(X)$ is smooth
%  \[\{U\in Gr^Q_\bfe(X):\lim_{t\to0} t\cdot U\in Gr^{\tilde Q}_{\tilde\bfe}(\tilde X)\}\]
%\end{corollary}

\begin{question}
  What are the ranks of the bundles over the fixed point set?  Poincar\'e polynomials?
\end{question}

@@

Using this notation, we obtain the following lemma:
\begin{lemma}
  \label{le:preprojective lifts}
  \begin{enumerate}
    \item $\Hom_Q(P_{m-1},P_m)\cong\bigoplus_{i=1}^n \Hom_{\tilde Q}(\tilde P_{m-1,i},\tilde P_m)$, where each $\Hom_{\tilde Q}(\tilde P_{m-1,i},\tilde P_m)$ is one-dimensional;
   % \item For any proper subset $\{i_1,\ldots,i_k\}\subset\{1,\ldots,n\}$, there exists a short exact sequence
   %   \[0\longrightarrow \tilde P_{m-1,i_1}\oplus\cdots\oplus\tilde P_{m-1,i_k}\longrightarrow\tilde P_m\longrightarrow \tilde P_m^{i_1,\ldots,i_k}\longrightarrow 0;\]
    \item The representations $\tilde P_{m-1,i}$ are pairwise orthogonal;
    \item All nontrivial proper subrepresentations $U$ of $\tilde P_m^{(i_1,\ldots,i_k)}$ are preprojective.
  \end{enumerate}
\end{lemma}
\begin{proof}
The first statement follows from Theorem \ref{covering}. The second part follows for instance recursively by applying BGP-reflections.

@@

\end{proof}
If we choose subset $\{i_1,\ldots,i_k\}=\{1,\ldots,k\}$, we denote the cokernel simply by $\tilde P_m^{(k)}$ and $\tilde P_m^{(0)}=\tilde P_m$ respectively.

@@

\end{proof}
The short exact sequence $\ses{\tilde P_{m,k}}{\tilde P_{m+1}^{(k-1)}}{\tilde P_{m+1}^{(k)}}$ induces a map between quiver Grassmannians
$$\Psi_1:Gr^{\tilde Q}_\tbfe(\tilde P_{m+1}^{(k-1)})\to\bigsqcup_{\tbff+\tbfg=\tbfe} Gr^{\tilde Q}_\tbff( \tilde P_{m,k})\times Gr^{\tilde Q}_\tbfg(\tilde P_{m+1}^{(k)}).$$
Let $P(m,k):=\bigoplus_{\substack{1\leq j\leq n\\j\neq k}}\tilde P_{m-1,(k,j)}\oplus \bigoplus_{j=1}^k\tilde P_{m-2,(k,k,j)}$.

@@

\begin{theorem}
  Every quiver Grassmannian of a preprojective or preinjective representation of $K(n)$ and $\widetilde{K(n)}$ has a cell decomposition.
\end{theorem}
\begin{proof}
This follows by combining the results of Section 2 and Proposition \ref{fibers}.
\end{proof}
\begin{question}
  cells of $Gr^{\tilde Q}_\tbfe(\tilde P_m)$ are in one-to-one correspondence with certain tuples of subgraphs for smaller $\tilde P_\ell^{i_1,\ldots,i_k}$
\end{question}

@@

\bibitem{gab} Gabriel, P.: The universal cover of a finite-dimensional algebra. Representations of algebras. Lecture Notes in Mathematics {\bf 903}, 68-105 (1981).
\bibitem{hr} Happel, D., Ringel, C.M.: Tilted Algebras. Transactions of the American Mathematical Society {\bf 274}, no.2, 399-443 (1982).
\bibitem{llz} K. Lee, L. Li, A. Zelevinsky: Greedy elements in rank 2 cluster algebras. Selecta Math. \textbf{20} (2014) pp.~57--82.
\bibitem{rin} Ringel, C.M.: Reflection functors for hereditary algebras. Journal of the London Mathematical Society (2) {\bf 21}, no. 3, 465-479 (1980).
%\bibitem{mats} Matsumura, H.: Commutative ring theory. 
\bibitem{rupel} Rupel, D.: Rank Two Non-Commutative Laurent Phenomenon and Pseudo-Positivity, arXiv.

>>>>

\newcommand{\bfe}{\mathbf{e}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\bfg}{\mathbf{g}}
\newcommand{\bfi}{\mathbf{i}}
\newcommand{\bfh}{\mathbf{h}}
\newcommand{\tbfe}{{\tilde\bfe}}
\newcommand{\tbff}{{\tilde\bff}}
\newcommand{\tbfg}{{\tilde\bfg}}
\newcommand{\tbfh}{{\tilde\bfh}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\cC}{\mathcal{C}}
\newcommand{\cG}{\mathcal{G}}
\newcommand{\cP}{\mathcal{P}}
\newcommand{\ui}{\underline i}
\newcommand{\uj}{\underline j}
\newcommand\udim{{\underline{\dim}\, }}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Rep}{\mathrm{Rep}}
\newcommand{\rep}{\operatorname{rep}}
\newcommand{\Gr}{\mathrm{Gr}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\kk}{\Bbbk}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\NN}{\mathbb{N}}

\newcommand{\Ext}{\operatorname{Ext}}
\newcommand{\Hom}{\operatorname{Hom}}
\newcommand{\im}{\operatorname{im}}
\newcommand{\codim}{\operatorname{codim}}
\newcommand{\into}{\hookrightarrow}
\newcommand{\onto}{\to\!\!\!\!\!\to}
\newcommand{\ses}[3]{0\rightarrow #1\rightarrow #2\rightarrow#3\rightarrow 0}
\newcommand{\Sc}[2]{\langle #1,#2\rangle}
\newcommand{\spn}{\operatorname{span}}
\setcounter{MaxMatrixCols}{20}

@@

\begin{theorem}Let $X$ be a representation which can be lifted to $\hat Q$ and choose $d:\mathrm{supp}(X)\to \ZZ$ such that it satisfies the degree condition for $X$. Then the torus action of $T$ on $\bigoplus_{q\in Q_0}X_q$ induced by $t.x_{(q,\chi)}=t^{d(q,\chi)}x_{(q,\chi)}$ for $x_{(q,\chi)}\in X_{(q,\chi)}$ induces
 a torus action on $Gr_\bfe^Q(X)$ such that
  $$Gr^Q_\bfe(X)^T\cong\bigsqcup_{\hat\bfe} Gr^{\hat Q}_{\hat\bfe}(\hat X)$$ where $\hat\bfe$ runs through all dimension vectors compatible with $\bfe$.
\end{theorem}
\begin{proof}
A representation $U\in\Gr_{\bfe}(X)$ is a torus fixed point if and only if $t.U=U $ for all $t\in\C^\ast$, i.e. $t.U_q=U_q$ for all $q\in Q_0$, $t\in\C^\ast$. Thus, apart from being a subrepresentation of $X$, it is a fixed point of the torus actions on the usual Grassmannians $\Gr_{\bfe_q}(X_q)$ induced by the decompositions $X_q=\bigoplus_{\chi\in A_Q} X_{q,\chi}$. A subrepresentation $U\subset X$ can be lifted to the universal abelian covering if and only if
\[X_q\cap U_q=\bigoplus_{\chi\in A_Q} X_{(q,\chi)}\cap U_{(q,\chi)}=\bigoplus_{\chi\in A_Q}U_{(q,\chi)}.\]
Thus we can apply Lemma \ref{usualGrass} to obtain the claim.
\end{proof}
The next step is to iterate the torus action keeping in mind the following idea: every representation, which lifts to the universal covering quiver, also lifts to the universal abelian covering quiver and also to the iterated universal abelian covering quivers, i.e. to $\hat Q^{(k)}:=\widehat{\hat Q^{(k-1)}}$ with $\hat Q=\hat Q^{(1)}$. Now it is straightforward to check that there exist natural surjective morphisms $f_k:\tilde Q\to \hat Q^{(k)}$ which become injective on finite subquivers if $k\gg 0$, see also \cite[Section 3.4]{wei}. Since the support of $X$ is finite as a representation of $\tilde Q$, we can find $k\geq 1$ such that the full subquiver with vertices $\mathrm{supp}(X)\subseteq \hat Q^{(k)}_0$ is a tree. Thus, if $\hat X^{(k)}$ is the lift of $X$ to $\hat Q^{(k)}$, there exists a torus action on the vector spaces $\hat X^{(k)}_q$ which induce a torus action on the quiver Grassmannian $Gr_{\bfe}^{\hat Q^{(k)}}(\hat X^{(k)})$. If we denote the iterated torus fixed points by $Gr^Q_\bfe(X)^{T^{(k)}}$. We obtain:
\begin{corollary}
Let $X$ be a representation which can be lifted to $\tilde Q$. Then there exists an iterated torus action such that
  $$Gr^Q_\bfe(X)^{T^{(k)}}\cong \bigsqcup_{\hat\bfe^{(k-1)}} Gr_{\hat\bfe^{(k-1)}}^{\hat Q^{(k-1)}}(\hat X^{(k-1)})^T\cong \bigsqcup_{\hat\bfe^{(k)}} Gr^{\hat Q}_{\hat\bfe^{(k)}}(\hat X^{(k)})\cong \bigsqcup_{\tilde\bfe} Gr^{\tilde Q}_{\tilde\bfe}(\tilde X)$$ where $\hat\bfe^{(k-1)},\,\hat\bfe^{(k)},\,\tilde\bfe$ run through all dimension vectors compatible with $\bfe$.
\end{corollary}
%\begin{corollary}
%  affine bundles over $Gr^{\tilde Q}_{\tilde\bfe}(\tilde X)$ if $Gr^Q_\bfe(X)$ is smooth
%  \[\{U\in Gr^Q_\bfe(X):\lim_{t\to0} t\cdot U\in Gr^{\tilde Q}_{\tilde\bfe}(\tilde X)\}\]
%\end{corollary}
Define the $F$-polynomial of a representation $X$ by 
$$F_X=\sum_{\bfe\in\NN^{Q_0}}\chi(\Gr_\bfe(X))x^e\in\ZZ[x_q\mid q\in Q_0].$$

\begin{corollary}\label{fpoly}
Let $X$ be a representation which can be lifted to the universal covering quiver.
\begin{enumerate}
\item 
If  $\Gr_\bfe(X)$ is smooth and $\Gr^{\tilde Q}_{\tilde \tbfe}(\tilde X)$ has a cell decomposition into affine spaces, then $\Gr_\bfe(X)$ has a cell decomposition into affine spaces.
\item We have $F_X=SF_{\tilde X}$ where $SF_{\tilde X}$ is obtained from $F_{\tilde X}$ by applying $S:\ZZ[x_{(q,w)}\mid q\in Q_0,w\in W_Q]\to\ZZ[x_q\mid q\in Q_0]$ given by $S(x_{(q,w)})=x_q$ for all $q\in Q_0$ and $w\in W_Q$.
\end{enumerate} 
\end{corollary}
An important special case for this is the case of exceptional representations. In this case the quiver Grassmannians $\Gr_\tbfe(X)$ are smooth by \cite[Corollary 4]{cr}. Moreover, every exceptional representation is a tree module by \cite{rin1} which means that it can be lifted to the universal covering.

@@

Using this notation, we obtain the following lemma:
\begin{lemma}
  \label{le:preprojective lifts}
  \begin{enumerate}
    \item $\Hom_Q(P_{m-1},P_m)\cong\bigoplus_{i=1}^n \Hom_{\tilde Q}(\tilde P_{m-1,i},\tilde P_m)$, where each $\Hom_{\tilde Q}(\tilde P_{m-1,i},\tilde P_m)$ is one-dimensional;
   % \item For any proper subset $\{i_1,\ldots,i_k\}\subset\{1,\ldots,n\}$, there exists a short exact sequence
   %   \[0\longrightarrow \tilde P_{m-1,i_1}\oplus\cdots\oplus\tilde P_{m-1,i_k}\longrightarrow\tilde P_m\longrightarrow \tilde P_m^{i_1,\ldots,i_k}\longrightarrow 0;\]
    \item The representations $\tilde P_{m-1,i}$ are pairwise orthogonal;
    \item All nontrivial proper subrepresentations $U$ of $\tilde P_m^{(i_1,\ldots,i_k)}$ are preprojective.

  \end{enumerate}
\end{lemma}
\begin{proof}
The first statement follows from Theorem \ref{covering}. The second part follows for instance recursively by applying BGP-reflections.

@@

%Fix $\ui=(i_1,\ldots,i_k)$ and $\uj=(j_1,\ldots,j_l)$ with $i$

If we choose subset $\{i_1,\ldots,i_k\}=\{1,\ldots,k\}$, we denote the cokernel simply by $\tilde P_m^{(k)}$ and $\tilde P_m^{(0)}=\tilde P_m$ respectively.

@@

The short exact sequence $\ses{\tilde P_{m,k}}{\tilde P_{m+1}^{(k-1)}}{\tilde P_{m+1}^{(k)}}$ induces a map between quiver Grassmannians
$$\Psi_1:Gr^{\tilde Q}_\tbfe(\tilde P_{m+1}^{(k-1)})\to\bigsqcup_{\tbff+\tbfg=\tbfe} Gr^{\tilde Q}_\tbff( \tilde P_{m,k})\times Gr^{\tilde Q}_\tbfg(\tilde P_{m+1}^{(k)}).$$
Let $P(m,k):=\bigoplus_{\substack{1\leq j\leq n\\j\neq k}}\tilde P_{m-1,(k,j)}\oplus \bigoplus_{j=1}^k\tilde P_{m-2,(k,k,j)}$.

@@

\begin{theorem}\label{celldec}
  Every quiver Grassmannian of a preprojective or preinjective representation of $K(n)$ and $\widetilde{K(n)}$ has a cell decomposition.
\end{theorem}
\begin{proof}
This follows by combining the results of Section 2 and Proposition \ref{fibers}.
\end{proof}

As the $F$-polynomial of truncated preprojective representations only depend on the dimension vector, we may denote them by $F_{\udim P_m{(k)}}$. The description of the non-empty fibres in Proposition \ref{fibers} together with Corollary \ref{fpoly} and Theorem \ref{celldec} yield the following:
\begin{corollary}
For $m\geq 2$ and $0\leq k\leq n-2$, we have 
$$F_{\udim P_m^{(k)}}=F_{\udim P_m^{(k+1)}}F_{\udim P_{m-1}}-x^{\udim P_{m}^{(k)}}F_{\udim P_{m-2}^{(k)}}.$$
\end{corollary}

@@

\bibitem{gab} Gabriel, P.: The universal cover of a finite-dimensional algebra. Representations of algebras. Lecture Notes in Mathematics {\bf 903}, 68-105 (1981).
\bibitem{hr} Happel, D., Ringel, C.M.: Tilted Algebras. Transactions of the American Mathematical Society {\bf 274}, no.2, 399-443 (1982).
\bibitem{llz} K. Lee, L. Li, A. Zelevinsky: Greedy elements in rank 2 cluster algebras. Selecta Math. \textbf{20} (2014) pp.~57--82.
\bibitem{rin1} Claus Michael Ringel. Exceptional modules are tree modules. \textit{Linear algebra and its Applications}, 275/276:471-493, 1998.
\bibitem{rin} Ringel, C.M.: Reflection functors for hereditary algebras. Journal of the London Mathematical Society (2) {\bf 21}, no. 3, 465-479 (1980).
%\bibitem{mats} Matsumura, H.: Commutative ring theory. 
\bibitem{rupel} Rupel, D.: Rank Two Non-Commutative Laurent Phenomenon and Pseudo-Positivity, arXiv.
